Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
consisntent of section 4 wrt section 3

\begin{definition}[Index scheme]
An \emph{index scheme} is a vector of size $n$ (an $n$-vector)
with values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$. 
Values in $\mathbb{Z}$ represent offsets 
of a neighbourhood index (indexing
expressions of the form $i + a$ where $i$ is an induction variable
and $a$ is a constant offset, Definition~\ref{def:neighbour},
p.~\pageref{def:neighbour}). The additional value
$\infty$ represents any indexing behaviour--- it is a 
wildcard.
Throughout $u, v, w$ will range over index schemes.
\end{definition}

@@

For each assignment statement whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed which
maps array variables to a set of vectors representing array
subscripts.  This finite map contains all array subscript expressions
which flow to this statement. More formally, a function
$\textsf{analyse}$ is applied to each statement in a program with the
following clause:
%
\begin{align*}
\begin{array}{lr}
\textsf{analyse}(v(\overline{e_1}) = e_2)
 := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{\textsf{convert}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(v')\}}
\end{array}
\end{align*}
%
That is, we focus on assignments to an array subscript where each
index indexing expression in $\bar{e}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed from each array variable
in this flow set to a representation of the subscripts, computed
with \textsf{convert}.

The \textsf{convert} function maps subscripts to a column vector
representation with values drawn from $\mathbb{Z}_\top = \mathbb{Z} \cup \{\top\}$
where $\mathbb{Z}$ represents the offset of a neighbourhood index
and $\top$ represents any non-neighbourhood subscripts, defined:
%
\begin{align*}
& \textsf{convert}((e_1, \ldots, e_n)) = [\textsf{conv}(e_1), \ldots,
  \textsf{conv}(e_n)] \\
& \textsf{conv}(e) \begin{cases}
\pm a & \neigh(e) \wedge \, e \equiv i \pm a \\
\top & \textit{otherwise}
\end{cases}
\end{align*}
%
For our example, the \textsf{analyse} function matches on
line 5, with the following set for $\textsf{flowTo}(a(i, j) + x +
  y)$:
%
\begin{align*}
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
\end{align*}
Subsequently the result of \textsf{analyse} on line 5 yields the map:
\begin{align*}
\texttt{a} \mapsto \{\vtwo{-1}{0}, \vtwo{1}{0},
          \vtwo{0}{-1}, \vtwo{0}{1}, \vtwo{0}{0}\}
\end{align*}
%

@@

Let $M$ range over the finite maps generated by \textsf{analyse}.  For
each $v \in \mathsf{dom}(M)$, the algorithm then constructs a group
of regions which cover all contiguous groups of relative indices
(from $M(v)$) in the $n$-dimensional space.

Informally, the procedure proceeds as follows. First, relative
indices are turned into unit regions in $n$-dimensions. For
example, for the five point stencil there are five $1 \times 1$ regions,
called \emph{spans}, which are essentially $0$-dimensional:
%
\begin{center}
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1,1) rectangle (1.5,1.5);
\draw[thick] (1.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (0.5,1) rectangle (1,1.5);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
\end{center}
%
Each dimension is traversed, coalescing consecutive spans
which vary in $m$-dimensions into coalesced spans which vary in
$m+1$-dimensions. In this example, the $0$-dimensional spans
 become $1$-dimensional spans (rows/columns):
%
\begin{align*}
\begin{array}{cc}
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (0.5,1) rectangle (2,1.5);
\draw[thick] (1,0.40) rectangle (1.5,0.90);
\draw[thick] (1,1.60) rectangle (1.5,2.10);
\end{tikzpicture}
\qquad
&
\qquad
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1.6,1) rectangle (2.1,1.5);
\draw[thick] (0.4,1) rectangle (0.9,1.5);
\draw[thick] (1,0.5) rectangle (1.5,2);
\end{tikzpicture}
\\
\textit{dimension 1} \qquad & \qquad \textit{dimension 2}
\end{array}
\end{align*}
%
Any spans that are contained with any other span are deleted,
leaving a minimal set of spans (which may overlap, but none of which
fully contains another).
\begin{center}
\begin{tikzpicture}
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1,0.5) rectangle (1.5,2);
\node at (2,1.22) {+};
\draw[thick] (2.5,1) rectangle (4,1.5);
%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
\end{tikzpicture}
\end{center}
%
The procedure is then iterated till a fixed point is reached. In this
example this is reached in the first step. These final spans
go on to form the basis of the spatial stencil specification, where
each span is a region product \texttt{*} and multiple spans are
combined with the region sum \texttt{+}.

\newcommand{\spanOp}{\textsf{spans}}
\paragraph{Span coalescing, formally}

Let $\vect{x}, \vect{y}, \vect{z}$ range over the column vectors of size
$n$ whose values are drawn from $\mathbb{Z}_\top$.
We write $\vect{x}_i$ for the $i$-th value of vector $\vect{x}$ (1
indexed) and $\vect{x}_{i:j}$ for the sub-vector from entry $i$ to
entry $j$ (inclusive).

\begin{definition}[Spans]
  A \emph{span} represents an $n$-dimensional box (\emph{hyper-rectangle}) as
  a pair of $n$-dimensional vectors (represented as a $2 \times n$
  matrix) giving the co-ordinates of the lower-bound vertex (first
  column) and the upper-bound vertex (second column). We use the
  notation $[\vect{x}^L \; \vect{x}^U]$ for such a span to mark out
  the lower and upper bound vectors.
\end{definition}

@@

\textbf{The Unified Model}\footnote{\url{http://www.metoffice.gov.uk/research/modelling-systems/unified-model}} is a weather forecasting and climate modelling tool developed by the Met Office in the United Kingdom. It is used by research organisations and meteorological services around the world. We use the development branch (trunk) of the model. The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model. We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.

@@

\textbf{Navier} is an implementation of a Navier-Stokes fluid dynamics computation~\cite{griebel1997numerical}.

>>>>

\begin{definition}[Index scheme]
An \emph{index scheme} is a vector of size $n$ (an $n$-vector)
with values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$. 
Values in $\mathbb{Z}$ represent offsets 
of a neighbourhood index (indexing
expressions of the form $i + a$ where $i$ is an induction variable
and $a$ is a constant offset, Definition~\ref{def:neighbour},
p.~\pageref{def:neighbour}). The additional value
$\infty$ represents any indexing behaviour--- it is a 
wildcard.
Throughout $u, v, w$ range over index schemes, and $u_i$ denotes
the $i^{th}$ element of $u$. 
\end{definition}

@@

For each assignment statement whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed which
maps array variables to a set of vectors representing array
subscripts.  This finite map contains all array subscript expressions
which flow to this statement. More formally, a function
$\textsf{analyse}$ is applied to each statement in a program with the
following clause:
%
\begin{align*}
\begin{array}{lr}
\textsf{analyse}(v(\overline{e_1}) = e_2)
 := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{\textsf{schematic}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(v')\}}
\end{array}
\end{align*}
%
That is, we focus on assignments to an array subscript where each
indexing expression in $\bar{e}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
in this flow a set of indexing schemes for its subscripts, computed
with \textsf{schematic}.

For our example, the \textsf{analyse} function matches on
line 5, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
  y})$:
%
\begin{align*}
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
\end{align*}
Subsequently the result of \textsf{analyse} on line 5 yields the map:
\begin{align*}
\texttt{a} \mapsto \{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\}
\end{align*}
%

@@

Let $M$ range over the finite maps generated by \textsf{analyse}.  For
each $v \in \mathsf{dom}(M)$, the algorithm then constructs a group
of regions (called \emph{spans}) which cover all contiguous groups of relative indices
(from $M(v)$) in the $n$-dimensional space.
Informally, indexing schemes are converted to pairs $u \mapsto (u,
u)$, defining the lower and upper bound of a unit span 
in $n$-dimensions. For
example, for the five point stencil there are five $1 \times 1$ spans.
%
\begin{center}
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1,1) rectangle (1.5,1.5);
\draw[thick] (1.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (0.5,1) rectangle (1,1.5);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
\end{center}
%
Each dimension is traversed, coalescing consecutive spans
which vary in $m$-dimensions into coalesced spans which vary in
$m+1$-dimensions. In this example, the $0$-dimensional spans
 become $1$-dimensional spans (rows/columns):
%
\begin{align*}
\begin{array}{cc}
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (0.5,1) rectangle (2,1.5);
\draw[thick] (1,0.40) rectangle (1.5,0.90);
\draw[thick] (1,1.60) rectangle (1.5,2.10);
\end{tikzpicture}
\qquad
&
\qquad
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1.6,1) rectangle (2.1,1.5);
\draw[thick] (0.4,1) rectangle (0.9,1.5);
\draw[thick] (1,0.5) rectangle (1.5,2);
\end{tikzpicture}
\\
\textit{dimension 1} \qquad & \qquad \textit{dimension 2}
\end{array}
\end{align*}
%
Any spans that are contained with any other span are deleted,
leaving a minimal set of spans (which may overlap, but none of which
fully contains another).
\begin{center}
\begin{tikzpicture}
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1,0.5) rectangle (1.5,2);
\node at (2,1.22) {+};
\draw[thick] (2.5,1) rectangle (4,1.5);
%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
\end{tikzpicture}
\end{center}
%
This procedure is iterated till a fixed point is reached. In this
example this is reached in the first step. The final spans
are used to construct stencil specification, where
each span is product \texttt{*} of regions and multiple spans are
combined with the region sum \texttt{+}.

\newcommand{\spanOp}{\textsf{spans}}

Let $\vect{x}, \vect{y}, \vect{z}$ range over the column vectors of size
$n$ whose values are drawn from $\domainVal{}$.
We write $\vect{x}_{i:j}$ for the sub-vector from entry $i$ to
entry $j$ (inclusive).

\begin{definition}[Spans]
  A \emph{span} represents an $n$-dimensional box (\emph{hyper-rectangle}) as
  a pair of $n$-dimensional vectors (represented as a $2 \times n$
  matrix) giving the co-ordinates of the lower-bound vertex (first
  column) and the upper-bound vertex (second column). We use the
  notation $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
  and $\vect{x}^U$ give the lower and upper bounds of the span.
\end{definition}

@@

\textbf{The Unified
  Model}\footnote{\url{http://www.metoffice.gov.uk/research/modelling-systems/unified-model}}
is a weather forecasting and climate modelling tool developed by the
Met Office in the United Kingdom. It is used by research organisations
and meteorological services around the world. We use the development
branch (trunk) of the model. The code based in closed source but
institutional licenses are available for research purposes. The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model. We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.

@@

\textbf{Navier} is a smaller projecter, giving a discrete
approximation to the two-dimensional Navier-Stokes fluid equations,
based on the book of~\citet{griebel1997numerical}.
